Image Caption Generation With CLIP+GPT-2 Model

### 2025-05-04 Latex_Paper/Paper.tex
update all

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{microtype}
% \usepackage{titlesec}
% \titleformat{\subsubsection}[runin]
%   {\normalfont\normalsize} 
%   {\thesubsubsection}      
%   {1em}                    
%   {}                     
%   []       
% \titlespacing*{\subsubsection}
%   {0pt}                       
%   {1.5ex plus 0.5ex minus 0.2ex} 
%   {0.8ex plus 0.2ex}                 
\usepackage{tikz}

@@

    \subsubsection{Data Set}

@@

This objective function aims to find the parameter set $\hat{\theta}$ that minimizes the total negative log-likelihood 
across the training dataset. During training, the model updates $\theta$ using optimization 
techniques such as gradient descent, guided by cross-validation to avoid overfitting and improve generalization.

\textbf{Training Details:} We train our model using these hyperparameter after cross-validation:
\begin{itemize}
  \item Prefix length: 40
  \item Transformer layers: 8
  \item Optimizer: Adam with a learning rate of $5 \times 10^{-5}$
  \item Number of epochs: 20
  \item Noise rate: 10\%
\end{itemize}

\subsection{Evaluation Protocol}
We evaluate the quality of learned representations by training a simple linear classifier on top of the
frozen embeddings.

\section{Experiments and Results}

\subsection{Dataset and Preprocessing}
We conduct experiments on:
\begin{itemize}
  \item \textbf{CIFAR-10:} A 10-class dataset with 60,000 images.
  \item \textbf{STL-10:} A larger dataset often used for unsupervised learning benchmarks.
\end{itemize}

\subsection{Baseline Comparisons}
We compare SimCLR embeddings with:
\begin{itemize}
  \item PCA-based dimensionality reduction ($d = 128$)
  \item Autoencoders trained on the same dataset
  \item Supervised ResNet-18 trained on CIFAR-10
\end{itemize}

\begin{table}[h]
\centering
\begin{tabular}{lcc}
\toprule
Method & CIFAR-10 Accuracy (\%) & STL-10 Accuracy (\%) \\
\midrule
Supervised ResNet-18 & 92.5 & 85.4 \\
PCA + kNN & 45.6 & 38.2 \\
Autoencoder + kNN & 55.3 & 49.6 \\
SimCLR (Ours) & 80.2 & 76.4 \\
\bottomrule
\end{tabular}
\caption{Comparison of representation learning methods. SimCLR significantly outperforms classical techniques.}
\end{table}

\subsection{Ablation Studies}
\textbf{Effect of Temperature $\tau$:} We analyze how different values of $\tau$ in the contrastive loss impact performance.

@@

\subsection{Key Findings}
\begin{itemize}
  \item SimCLR significantly outperforms PCA and autoencoders in feature learning.
  \item The choice of augmentations greatly affects performance.
  \item Higher temperature values in contrastive loss lead to better separation of features.
\end{itemize}

@@

    \bibitem{papineni2002bleu}
    Papineni, K., Roukos, S., Ward, T., \& Zhu, W. J. (2002). ``BLEU: a method for automatic evaluation of machine translation.'' ACL 2002.

@@

    \bibitem{chen2020simple}
    Ting Chen, Simon Kornblith, Mohammad Norouzi, Geoffrey Hinton. ``A Simple Framework for Contrastive Learning of Visual Representations.'' ICML 2020.

    \bibitem{hinton2006reducing}
    Geoffrey Hinton, Ruslan Salakhutdinov. ``Reducing the Dimensionality of Data with Neural Networks.'' Science, 2006.
    \end{thebibliography}

>>>>

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{microtype}
% \usepackage{titlesec}
% \titleformat{\subsubsection}[runin]
%   {\normalfont\normalsize} 
%   {\thesubsubsection}      
%   {1em}                    
%   {}                     
%   []       
% \titlespacing*{\subsubsection}
%   {0pt}                       
%   {1.5ex plus 0.5ex minus 0.2ex} 
%   {0.8ex plus 0.2ex}                 
\usepackage{tikz}

@@

    \subsubsection{DataSet}

@@

This objective function aims to find the parameter set $\hat{\theta}$ that minimizes the total negative log-likelihood 
across the training dataset. During training, the model updates $\theta$ using optimization 
techniques such as gradient descent, guided by cross-validation to avoid overfitting and improve generalization.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{f2.png}
    \caption{Training and validation loss over epochs}
    \end{figure}
\pagebreak
\section{Experiments and Results}
\subsection{Training Details}
We train our model using these hyperparameter after cross-validation:
\begin{itemize}
  \item Prefix length: 40
  \item Transformer layers: 8
  \item Number of epochs: 12
  \item Noise rate: 10\%
  \item Optimizer: Adam with a learning rate of $5 \times 10^{-5}$
\end{itemize}

We conduct experiments on:
\begin{itemize}
  \item \textbf{COCO Image Dataset:}

  Training set: 80, 000 images with 5 captions per image.

  Validation set: 2,000 images.

\end{itemize}

\subsection{Evaluation Metrics}
After completing model training, we need to evaluate the quality of generated captions to assess the effectiveness of our approach. Therefore, we adopt two widely used 
automatic evaluation metrics in the field of image captioning: BLEU and CIDEr.
\begin{itemize}
    \item \textbf{Bilingual Evaluation Understudy }

    BLEU is a baseline evaluation 
    metric of our model that measures the accuracy of n-gram overlap between predicted captions 
    and one or more real captions\cite{Papineni}. It emphasizes syntactic similarity and is particularly sensitive to exact word 
    matching and word order. However, this criterion lacks semantic awareness and often penalizes correct words due to different word position information.

    \item \textbf{Consensus-based Image Description Evaluation}
    To address the shortcomings of the above evaluation method, we also chose the CIDEr
     (Consensus-based Image Description Evaluation) \cite{vedantam2015cider}
     evaluation metric, which is designed specifically for image 
     captioning. It uses a TF-IDF weighted n-gram model to evaluate how well the 
     generated captions match the manually generated captions. This design increases the 
     weights of rare but informative words (e.g., nouns, verbs, adjectives, etc., which often contain unique image information) 
     while reducing the weights of function words such as articles (a, the), quantifiers (some, many), and prepositions (in, on, at).
\end{itemize}

\pagebreak
\subsection{Generation Results}
\textbf{Prediction Scores:}

We compute BLEU and CIDEr on our test set and compare the original captions' scores with our outputs:
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{f3.png}
    \caption{Model vs Human Performance on Image Captioning}
    \end{figure}
\begin{itemize}
    \item Human captions show better performance on longer word sequences, indicating more natural grammatical structure.
    \item The CIDEr score shows our model can achieve 94.4\% of human-level consensus quality.
\end{itemize}
\pagebreak
\textbf{Web-based Deployment for Real-time Captioning}

We design and deploy a website to realize our model
function. The website is able to upload images and load the pre-trained model to generate 
captions within 10 seconds.

\begin{figure}[h]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{F4.png}
        \caption{Example 1}
        \label{fig:upload}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.43\textwidth}
        \centering
        \includegraphics[width=\textwidth]{F5.png}
        \caption{Example 2}
        \label{fig:caption}
    \end{subfigure}
    \caption{Web demo of our deployed model: From image upload to automatic caption generation}
    \label{fig:web_demo}
\end{figure}

@@

\subsection{Key Findings}
\begin{itemize}
  \item Automated organization of visual content:

  Our model efficiently manages unlabeled images, reducing the need for manual effort.  
  \item Effective CLIP-to-GPT-2 Mapping:

  Our multi-head attention mechanism bridges visual and textual modalities for high-quality captions.
  \item Noise Injection Boosts Generalization:

  Introducing noise during training boosts generalization and helps create diverse, accurate captions.
\end{itemize}

@@

    \bibitem{Papineni}
    Papineni, K., Roukos, S., Ward, T., \& Zhu, W. J. (2002). ``BLEU: a method for automatic evaluation of machine translation.'' ACL 2002.

@@

    \end{thebibliography}
